---
title: "Exam"
author: "Laura Nyrup Mogensen"
date: "7/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
library(spatstat)
library(RandomFields)

```

## Cox Processes
Simulation of a Matérn cluster Process:
```{r}
# Parameters
kappa1 = 15
scale1 = 0.09
mu1 = 6.5

X = rMatClust(kappa1, scale1, mu1) # point process

# Visualization of point process
plot(X, "Matérn Cluster Process")

npoints(X)
# there is 66 points
```

Simulation of a Thomas Process:
```{r}
# Parameters
kappa2 = 23.5
scale2 = 0.05
mu2 = 2.6

Y = rThomas(kappa2, scale2, mu2) # Point Process

# Visualization of the point process
plot(Y, "Thomas Proces")

npoints(Y)
# there is 68 points
```

Simulation of arbitrary Neyman-Scott Processes:
```{r}
# Parameters
kappa3 = 9
rmax = 3


thin <- function(x,y,r,x0,y0)
{
  ifelse((x-x0)^2 + (y-y0)^2 < r^2, 0, 1)
}

nclust <- function(x0, y0, r1, r2, beta)
{
  num = rpois(1, beta)
  xc = runifdisc(num, r2, centre=c(x0, y0))
  xc = rthin(xc, thin, r=r1, x0=x0, y0=y0)
}

Z <- rNeymanScott(kappa3, rmax, nclust, r1=0.2, r2=5, beta=2)

# Visualization of point process
plot(Z, "Neyman-Scott Process")

npoints(Z)
# there is 12 points
```

Visualization of the `redwood` data set:
```{r}
plot(redwood)
npoints(redwood)
# there is 62 points
```

```{r}
rm(list = ls())
```

## Log Point Process
```{r}
mu = 5
var = 0.2
scale = 0.1
```

Simulation of a Gaussian Log Point Process
```{r}
X1 = rLGCP("gauss", mu, var, scale) # delta = 2
Random_field1 = attr(X1, "Lambda"); Random_field1
plot(Random_field1, "Gaussian covariance")
points(X1)
```

Simulation of a exponential Log Point Process:
```{r}
X2 = rLGCP("exp", mu, var, scale)   # delta = 1
Random_field2 = attr(X2, "Lambda"); Random_field2
plot(Random_field2, "Exponential covariance")
points(X2)
```

Simulation of Power Exponential family
```{r}
mu = 5
var = 0.2
scale = 0.1
alpha = 0.1

X3 = rLGCP("stable", mu, var, scale, alpha)
Random_field3 = attr(X3, "Lambda"); Random_field3
plot(Random_field3, "Power exponential covariance")
points(X3)
```

```{r}
rm(list = ls())
```

## Poisson Point Process
Simulation of a homogen Poisson Process with intensity `lambda = 100` on the unit square in R2.
```{r}
lambda = 100
Homogen_Poisson = rpoispp(lambda, lmax=NULL, owin(c(0,1),c(0,1))) # Laver en Poisson punkt proces
plot(Homogen_Poisson)
```

Simulation of a inhomogen Poisson Process with intensity `lambda = 100` on the square [-1, 1] × [-1, 1] in R2.
```{r}
lambda = function(x,y,a,b) {a*dnorm(x,0,b)*dnorm(y,0,b)}

Inhomogen_Poisson = rpoispp(lambda, win=owin(c(-1,1),c(-1,1)),a=100,b=0.3); Inhomogen_Poisson
plot(Inhomogen_Poisson)
```

Simulation of a thinning of a Poisson Point Process:
```{r}
lambda = 300
w = owin(c(-1,1),c(-1,1))
X = rpoispp(300, win=w)
plot(X, "Poisson punkt proces")

p = function(x,y) 
{ 
  ifelse(x^2 + y^2 < 0.2, 1, 0.2) 
}

Y = rthin(X, p)           # Ny Poisson proces
plot(Y, "Udtyndet Poissen punkt proces")
# rthin bruger en punkt proces og en sandsynlighed for at beholde et punkt som input
```

```{r}
rm(list = ls())
```

## Thomas Point Process
For simulation of the Thomas Point Process the `redwood` data set will be use for the minimum contrast estimation.
```{r}
data(redwood)
summary(redwood) # Giver information om data
plot(redwood)
```

The summary statistics of the pair correlation function.
```{r}
env = envelope(redwood, fun = pcf, nsim = 39)
plot(env, ylim=c(0,5)) # Vi ser de teoretiske og observeret data for PCF
```

Estimation of the intensity with kernels-
```{r}
intensity = density.ppp(redwood, 0.05) # sigma er valg til at være 0.05
plot(intensity)
points(redwood)
```

The fitting model - the Thomas Process -- is performed with the K-function as a theoretical value.
```{r}
model = thomas.estK(redwood)
plot(model, main = c("Thomas proces", "fit ved minimum kontrast"))
model # Giver information om modellen:
```

Lets simulate the Thomas process.
```{r}
para = model$modelpar
Xsim = rThomas(kappa=para[1], sigma=para[2], mu=para[3])
plot(Xsim)
```

Lets make a model check and visual comparison of the simulation and the data.
```{r}
par(mfrow=c(1,2))
Xsim = rThomas(kappa=para[1], sigma=para[2], mu=para[3])
plot(redwood)
plot(Xsim)
par(mfrow=c(1,1))
```

```{r}
rm(list = ls())
```

## Summary Statistics
I will be using the data set `redwood` and `cells`.
```{r}
plot(redwood)
plot(cells)
```

Lets look at their K, L, F, G og J-functions
```{r}
K1 = Kest(redwood)
L1 = Lest(redwood)
F1 = Fest(redwood)
G1 = Gest(redwood)
J1 = Jest(redwood)

K2 = Kest(cells)
L2 = Lest(cells)
F2 = Fest(cells)
G2 = Gest(cells)
J2 = Jest(cells)

par(mfrow=c(1,2))
plot(K1); plot(K2)
plot(L1); plot(L2)
plot(F1); plot(F2)
plot(G1); plot(G2)
plot(J1); plot(J2)
```

`redwood` lays above the Poisson process which mean that `redwood` is clustered. `cells` laysr below the Poisson process which mean that `cells` is regular.

If L is plotted as L(r) with r as it default.
```{r}
plot(L1,.-r~r); plot(L2,.-r~r)
```

A point-wise envelopes based on 39 simulations for a homogen Poisson proces
```{r}
e1 = envelope(redwood, Lest, nsim=39) 
e2 = envelope(cells, Lest, nsim=39)
plot(e1, .-r~r); plot(e2, .-r~r)
```

Same conlusion is made for `redwood` and `cells`.

```{r}
rm(list = ls())
```

## MCMC
Defining a MCMC for Strauss process on a unit square
```{r}
nr = 10000    # Antal gentagelser i MCMC
nv = 0        # Printer hvor lang algoritmen er nået hver nv'th step. 0 = ingen print

model1 = list(cif="strauss", par=list(beta=200, gamma=0.2, r=0.07), w=c(0,1,0,1))
model2 = list(cif="strauss", par=list(beta=100, gamma=0.2, r=0.07), w=c(0,1,0,1))
model3 = list(cif="strauss", par=list(beta=200, gamma=0.2, r=1), w=c(0,1,0,1))
model4 = list(cif="strauss", par=list(beta=200, gamma=0.2, r=0.01), w=c(0,1,0,1))
model5 = list(cif="strauss", par=list(beta=200, gamma=0.1, r=0.07), w=c(0,1,0,1))
model6 = list(cif="strauss", par=list(beta=200, gamma=1, r=0.07), w=c(0,1,0,1))

start = list(n.start=80)
control1 = list(nrep=nr, nverb=nv)

X1 = rmh(model=model1, start=start, control=control1)
X2 = rmh(model=model2, start=start, control=control1)
X3 = rmh(model=model3, start=start, control=control1)
X4 = rmh(model=model4, start=start, control=control1)
X5 = rmh(model=model5, start=start, control=control1)
X6 = rmh(model=model6, start=start, control=control1)

npoints(X1); plot(X1, "Strauss proces, beta = 200")
npoints(X2); plot(X2, "Strauss proces, beta = 100")
npoints(X3); plot(X3, "Strauss proces, r = 1")
npoints(X4); plot(X4, "Strauss proces, r = 0.01")
npoints(X5); plot(X5, "Strauss proces, gamma = 0.1")
npoints(X6); plot(X6, "Strauss proces, gamma = 1")
```

```{r}
rm(list = ls())
```
