---
title: "Data Visualization with R"
author: "Laura Nyrup Mogensen"
date: "5/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
With use of the `mpg` data frame I will go through the different athestics of visualize data in RStudio with use of the `tidyverse` library which include the `ggplot2` package.

```{r}
#install.packages("tidyverse") 

library(tidyverse)
#> ── Attaching packages ─────────────────────────────────────── tidyverse 1.3.0 ──
#> ✔ ggplot2 3.3.0     ✔ purrr   0.3.4
#> ✔ tibble  3.0.1     ✔ dplyr   0.8.5
#> ✔ tidyr   1.0.3     ✔ stringr 1.4.0
#> ✔ readr   1.3.1     ✔ forcats 0.5.0

#tidyverse_update() # Update if necessary
```

```{r}
mpg  # data frame
#?mpg # information about the data frame
```

Within the `mgp` data frame there are 234 rows and 11 columns. The 11 columns are the variables of the data frame:

* manufacturer: the manufacturer name
* model: the models name
* displ: the engine displacement, in litre
* year: year of manufacture
* cyl: number of cylinders
* tran: type of transmission
* drv: the type of driven train
* cty: city miles per gallon
* hwy: highway miles per gallon
* fl: fuel type
* class: 'type' of car

## Geom_point Plots
Lets investigate the continuous variables `displ` and `hwy`
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy))
```

The plot illustrate a negative relationship between `displ` and `hwy` as the `hwy` decrease when `displ` increase. From this plot we can see that cars with bigger engines use more fuel than cars with smaller engines.

### Categorical Variables
For data type which are objects it is also possible to visualize those together with the continuous data. This can be done by using the `color`, `size`, `alpha`or `shape` in the `aes` function of the `geom_point` plot function
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, color = class))
```

The `size` function is not advised for discrete variables.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, size = class))
```

```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, alpha = class))
```

For the `shape` the `ggplot2` can only visualize 6 different shapes and if there are more than 6 categorical values they will not be visualized.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy, shape = class))
```

Another way to visualize categorical variables is to use the `facet` functions. `facet_wrap()` need a formula (`~`) which is name of the data structure and the variable name.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_wrap(~ class, nrow = 2)
```

If you want to plot two categorical variables you need to use the `facet_grid()` function.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) + 
  facet_grid(drv ~ cyl)
```

## Geom_smoot Plots
The `geom` plot function have several syntax for different type of visualisations. The one I will be showing is the `geom_smooth()`.
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

To remove the transparency field around the line on should use the `se = FALSE` in the `geom_smooth()` function.
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy), se=FALSE)
```

There are several ways to visualize the `mpg` variables within the `geom_smooth()` function.
```{r}
ggplot(data = mpg) + 
  geom_smooth(mapping = aes(x = displ, y = hwy, linetype = drv))
```

```{r}
ggplot(data = mpg) +
  geom_smooth(mapping = aes(x = displ, y = hwy, group = drv))
```

```{r}
ggplot(data = mpg) +
  geom_smooth(
    mapping = aes(x = displ, y = hwy, color = drv),
    show.legend = FALSE
  )
```

## Two Plots in One Plot
It is also possible to display two different plot in the same plot by multiplying the two `geom` functions.
```{r}
ggplot(data = mpg) + 
  geom_point(mapping = aes(x = displ, y = hwy)) +
  geom_smooth(mapping = aes(x = displ, y = hwy))
```

To make sure that the plots used the same x- and y-axis and not making any error you can write define the axis with the data assignment.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point() + 
  geom_smooth()
```
Or like this where some other definitions to the plot are made which not necessary are needed for both plots.
```{r}
ggplot(data = mpg, mapping = aes(x = displ, y = hwy)) + 
  geom_point(mapping = aes(color = class)) + 
  geom_smooth()
```

## Geom_bar Plots
For illustrating the `geom_bar()` plot function we will be need a new data frame to use. In the ´ggplot2´ package the `diamonds` data frame fit to this illustration. Within the `diamonds` data frame there are 53940 rows and 10 columns. The 10 columns are the variables of the data frame:

* price: price in US dollars
* carat: weight of the diamond
* cut: quality of the cut
* color: diamond color
* clarity: a measurement of how clear the diamond is
* x: lenght in mm
* y: width in mm
* z: depth in mm
* depth: total depthm percentage
* table: width of top of diamond relatice to widest point

Lets visualize the `cut` variable.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut))
```

The count of the plot is the number of point which fall within each bin. 

There several aesthetics that follows with the `geom_bar()` function, such as the `color`and `fill` in the `aes` function.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, colour = cut))
```

```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = cut))
```

The bars can be stack if you want to include another variable to the ´geom_bar()` plot.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity))
```

The stacking is performed automatically by the position adjustment specified by the `position` argument. If you don’t want a stacked bar chart, you can use one of three other options: `identity`, `dodge` or `fill`.

The `identity` argument will place each object exactly where they falls in the context of the graph. Not useful for bars beacuse it overlaps them.
```{r}
ggplot(data = diamonds, mapping = aes(x = cut, fill = clarity)) + 
  geom_bar(alpha = 1/5, position = "identity")
ggplot(data = diamonds, mapping = aes(x = cut, colour = clarity)) + 
  geom_bar(fill = NA, position = "identity")
```

The `fill` argument works like stacking the bars but it make all the stacked bars the same height.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "fill")
```

The `dodge` argument places overlapping objects directly beside on another and this make it easier to compare indicidual values.
```{r}
ggplot(data = diamonds) + 
  geom_bar(mapping = aes(x = cut, fill = clarity), position = "dodge")
```

