---
title: "Exploratory Data Analysis with R"
author: "Laura Nyrup Mogensen"
date: "7/9/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
```{r}
#install.packages("tidyverse") 

library(tidyverse)
#> ── Attaching packages ─────────────────────────────────────── tidyverse 1.3.0 ──
#> ✔ ggplot2 3.3.0     ✔ purrr   0.3.4
#> ✔ tibble  3.0.1     ✔ dplyr   0.8.5
#> ✔ tidyr   1.0.3     ✔ stringr 1.4.0
#> ✔ readr   1.3.1     ✔ forcats 0.5.0

#tidyverse_update() # Update if necessary
library(nycflights13)
```
With exploratory data analysis (EDA) there are two type of question that need to be answer to understand and investigate the data:

* What type of variation occurs within my variables?
* What type of covariation occurs between my variables?

To answer those questions one need to understand the following terms:

* A `variable` is a quantity, quality, or property that you can measure.
* A `value` is the state of a variable when you measure it. 
* An `observation` is a set of measurements made under similar conditions. An observation will contain several values, each associated with a different variable. 
* `Tabular data` is a set of values, each associated with a variable and an observation. Tabular data is tidy if each value is placed in its own “cell”, each variable in its own column, and each observation in its own row.

## Variation


## Patterns & Models
Patterns in your data provide clues about relationships. If a systematic relationship exists between two variables it will appear as a pattern in the data. If you spot a pattern, ask yourself:

* Could this pattern be due to coincidence (i.e. random chance)?
* How can you describe the relationship implied by the pattern?
* How strong is the relationship implied by the pattern?
* What other variables might affect the relationship?
* Does the relationship change if you look at individual subgroups of the data?

A scatterplot of `Old Faithful` eruption lengths versus the wait time between eruptions shows a pattern: longer wait times are associated with longer eruptions.
```{r}
ggplot(data = faithful) + 
  geom_point(mapping = aes(x = eruptions, y = waiting))
```






